### 2014-09-10 src/paper1/p1_discussion.tex
New siang command. Created custom style file.

After earlier indications that both the otoliths and body sensors contribute to the SBT \cite{clark1963, nelson1968}, Mittelstaedt \citeyear{mittelstaedt1997} made a quantitative assessment of their impact, using an ingeniously designed experiment. Subjects lay on their side in a horizontal centrifuge. The crux of the experiment was to vary the distance between the rotation axis and the interaural axis to equalize the opposite contributions from the otoliths and the body sensors so that the subject felt horizontal. By testing normal, paraplegic, and nephrectomized subjects, Mittelstaedt inferred how much each sensory system contributes to body-tilt perception. It was shown that, apart from the otoliths, also internal ``graviceptors'' in the trunk (such as the viscera) participate in the computation of the SBT. Later, some related studies provided evidence that the distribution of blood in the body also affects postural perception \cite{vaitl1997, vaitl2002}. According to Mittelstaedt \citeyear{mittelstaedt1998}, the weight of the somatic graviceptors to estimate horizontal body orientation in healthy subjects is \textapprox0.6 on average, with considerable intersubject variability. His estimate seems quite compatible with our $w_{BD}$ values at 90\mydegree tilt, which range between 0.35 (S5) and 0.93 (S6) (Fig. \ref{p1:fig5}).

@@

The architecture of the model, as far as the reference frame transformations and the sensory integration is concerned, follows entirely from the principles of Bayesian inference. However, to account for our major findings and intersubject differences, we made two less straightforward assumptions. First, to explain the increased variability in both tasks at 90\mydegree tilt, we allowed for the possibility that the otoliths become more noisy with increases in tilt. Second, we hypothesized that prior knowledge is used in the visual vertical but not in body-tilt perception. Can these assumptions be justified on physiological and rational grounds?

One reason to assume that otolith noise depends on tilt angle is based on the fact that the utricle contains considerably more hair cells than the saccule \cite{rosenhall1972, rosenhall1974}. Because the utricle is most sensitive to tilts of \textapprox0\textdegree, whereas the saccule is most sensitive at \textapprox90\mydegree tilt \cite{jaeger2008}, this may well cause the proposed increase of otolith noise with tilt angle \cite{tarnutzer2010}. A tilt-dependent noise level of the otoliths would also help to explain why the perturbing effect of roll-optokinetic stimulation on the SVV \cite{dichgans1974, fernandez1976} and on the SBT \cite{young1975} is more pronounced at larger tilt angles and why the SVV is more strongly influenced by residual canal signals at larger tilt angles, after prolonged roll rotations \cite{lorincz2008}.

@@

According to our model, statistically optimal performance requires the use of information from both direct and indirect pathways to estimate body and head orientation in space (Fig. \ref{p1:fig1}). Thus, if one of the sensory inputs is lost or severely disrupted, SBT and SVV performance will deteriorate, but should not completely break down due to their multimodality dependence. By setting the appropriate parameter values of the model to infinity, we simulated the model to predict SVV and SBT performance in two patient groups: bilateral vestibular patients (noise level of the otoliths set to infinity) and patients with somatosensory loss (noise level of the body sensors set to infinity). Figure \ref{p1:fig7} shows the results of these simulations.

@@

    \caption{Clinical implications of the model. The model simulates the SBT and SVV in a vestibular patient by raising the level of the otolith noise to infinity, keeping the other param- eters at the mean values of Table \ref{p1:tab1}. A somatosensory patient is modeled by setting the noise level of the body sensors to infinity. Solid lines, Patient predictions. Dashed lines, Prediction for normals.}
    \label{p1:fig7}
\end{SCfigure}

@@

The right column of Figure \ref{p1:fig7} depicts a simulation of the SBT and SVV in a patient with loss of somatosensory information (somatosensory patient). In this case, the SBT depends solely on otolith information mediated through the indirect pathway. Although this signal is still accurate, it is spoiled by the larger variability of the neck signal, needed to perform the appropriate reference frame transformation. The model also predicts larger errors in the SVV in these patients than in normal subjects. Although there are no reports containing measurements of bias and variance in the SBT and SVV, paraplegic patients show that along with the otoliths, internal body sensors also contribute to the SBT if lesions are below the 12th thoracic segment \cite{mittelstaedt1997}. This evidence supports the design of our model.

@@

Here we provide further explanation about the Bayesian computations underlying the SVV as expressed in Equations \ref{p1:eqn6} and \ref{p1:eqn10} in Materials and Methods. Figure 8 illustrates graphically that the variance of the posterior distribution in a single trial ($\sigma^2_{\tilde{H}S}$) is not simply the same as the variance in its peak location in multiple trials, $\sigma^2(\tilde{H}_S)$. In a single trial (Fig. \ref{p1:fig8}A-C), the optimal estimate of head tilt is based on the likelihood (Fig. \ref{p1:fig8}B, green curve) associated with the combined sensory input from the direct and the indirect pathway (Fig. \ref{p1:fig8}A, green line, $\hat{H}_S$) and the prior (Fig. \ref{p1:fig8}B, blue curve), by multiplication of the two probability distributions. The prior distribution is a Gaussian with mean $H_{SP}$ and variance $\sigma^2_{HSP}$. The peak of the resulting posterior distribution (Fig. \ref{p1:fig8}B, orange curve) is used as the optimal estimate of head tilt ($\tilde{H}_S$), given by the following:

@@

in which $\sigma_{HS}$ denotes the noise in the sensory signal, known to the observer, and $w_{HS}$ and $w_{HP}$ represent the relative weights of the sensory signal and the prior, respectively. Note that Equation \ref{p1:eqn13} is equivalent to Equation \ref{p1:eqn1} in \nameref{p1:sec:methods}. The variance of the posterior distribution in a single trial is given by the following:

@@

and is reflected by the width of the orange curve in Figure \ref{p1:fig8}B. Figure \ref{p1:fig8}D-F illustrates performance in multiple trials, in which the posterior distributions vary due to sensory noise ($\sigma_{HS}$), whereas the prior remains fixed. The variance of each posterior distribution is fixed and is given by Equation \ref{p1:eqn16}.

@@

That the variance of the peak locations across multiple trials, $\sigma^2(\tilde{H}_S)$, is smaller can be shown by applying the rules of noise propagation to Equation \ref{p1:eqn13}.

@@

which is equivalent to Equation \ref{p1:eqn10} in Materials and Methods. Corresponding panels G-I (Fig. \ref{p1:fig8}) illustrate the distribution of the sensory signals for a given tilt angle (green-shaded curve), the prior distribution (blue-shaded curve), and the optimal estimates (orange-shaded curve), respectively. Figure \ref{p1:fig8}I illustrates that the distribution of the optimal estimates across many trials has a lower variance than the posterior distribution in each single trial (Fig. \ref{p1:fig8}B), which follows from the comparison of Equations \ref{p1:eqn16} and \ref{p1:eqn17}, respectively.

>>>>

After earlier indications that both the otoliths and body sensors contribute to the SBT \cite{clark1963, clark1964, nelson1968}, Mittelstaedt \citeyear{mittelstaedt1997} made a quantitative assessment of their impact, using an ingeniously designed experiment. Subjects lay on their side in a horizontal centrifuge. The crux of the experiment was to vary the distance between the rotation axis and the interaural axis to equalize the opposite contributions from the otoliths and the body sensors so that the subject felt horizontal. By testing normal, paraplegic, and nephrectomized subjects, Mittelstaedt inferred how much each sensory system contributes to body-tilt perception. It was shown that, apart from the otoliths, also internal ``graviceptors'' in the trunk (such as the viscera) participate in the computation of the SBT. Later, some related studies provided evidence that the distribution of blood in the body also affects postural perception \cite{vaitl1997, vaitl2002}. According to Mittelstaedt \citeyear{mittelstaedt1998}, the weight of the somatic graviceptors to estimate horizontal body orientation in healthy subjects is \textapprox0.6 on average, with considerable intersubject variability. His estimate seems quite compatible with our $w_{BD}$ values at \siang{90} tilt, which range between 0.35 (S5) and 0.93 (S6) (\figref{p1:fig5}).

@@

The architecture of the model, as far as the reference frame transformations and the sensory integration is concerned, follows entirely from the principles of Bayesian inference. However, to account for our major findings and intersubject differences, we made two less straightforward assumptions. First, to explain the increased variability in both tasks at \siang{90} tilt, we allowed for the possibility that the otoliths become more noisy with increases in tilt. Second, we hypothesized that prior knowledge is used in the visual vertical but not in body-tilt perception. Can these assumptions be justified on physiological and rational grounds?

One reason to assume that otolith noise depends on tilt angle is based on the fact that the utricle contains considerably more hair cells than the saccule \cite{rosenhall1972, rosenhall1974}. Because the utricle is most sensitive to tilts of \textapprox0\textdegree, whereas the saccule is most sensitive at \siang{\textapprox90} tilt \cite{jaeger2008}, this may well cause the proposed increase of otolith noise with tilt angle \cite{tarnutzer2010}. A tilt-dependent noise level of the otoliths would also help to explain why the perturbing effect of roll-optokinetic stimulation on the SVV \cite{dichgans1974, fernandez1976} and on the SBT \cite{young1975} is more pronounced at larger tilt angles and why the SVV is more strongly influenced by residual canal signals at larger tilt angles, after prolonged roll rotations \cite{lorincz2008}.

@@

According to our model, statistically optimal performance requires the use of information from both direct and indirect pathways to estimate body and head orientation in space (\figref{p1:fig1}). Thus, if one of the sensory inputs is lost or severely disrupted, SBT and SVV performance will deteriorate, but should not completely break down due to their multimodality dependence. By setting the appropriate parameter values of the model to infinity, we simulated the model to predict SVV and SBT performance in two patient groups: bilateral vestibular patients (noise level of the otoliths set to infinity) and patients with somatosensory loss (noise level of the body sensors set to infinity). \figref{p1:fig7} shows the results of these simulations.

@@

    \caption{Clinical implications of the model. The model simulates the SBT and SVV in a vestibular patient by raising the level of the otolith noise to infinity, keeping the other param- eters at the mean values of \tabref{p1:tab1}. A somatosensory patient is modeled by setting the noise level of the body sensors to infinity. Solid lines, Patient predictions. Dashed lines, Prediction for normals.}
    \label{p1:fig7}
\end{SCfigure}

@@

The right column of \figref{p1:fig7} depicts a simulation of the SBT and SVV in a patient with loss of somatosensory information (somatosensory patient). In this case, the SBT depends solely on otolith information mediated through the indirect pathway. Although this signal is still accurate, it is spoiled by the larger variability of the neck signal, needed to perform the appropriate reference frame transformation. The model also predicts larger errors in the SVV in these patients than in normal subjects. Although there are no reports containing measurements of bias and variance in the SBT and SVV, paraplegic patients show that along with the otoliths, internal body sensors also contribute to the SBT if lesions are below the 12th thoracic segment \cite{mittelstaedt1997}. This evidence supports the design of our model.

@@

Here we provide further explanation about the Bayesian computations underlying the SVV as expressed in \eqnref[ and ]{p1:eqn6,p1:eqn10} in Materials and Methods. \figref{p1:fig8} illustrates graphically that the variance of the posterior distribution in a single trial ($\sigma^2_{\tilde{H}S}$) is not simply the same as the variance in its peak location in multiple trials, $\sigma^2(\tilde{H}_S)$. In a single trial (\figref{p1:fig8}A-C), the optimal estimate of head tilt is based on the likelihood (\figref{p1:fig8}B, green curve) associated with the combined sensory input from the direct and the indirect pathway (\figref{p1:fig8}A, green line, $\hat{H}_S$) and the prior (\figref{p1:fig8}B, blue curve), by multiplication of the two probability distributions. The prior distribution is a Gaussian with mean $H_{SP}$ and variance $\sigma^2_{HSP}$. The peak of the resulting posterior distribution (\figref{p1:fig8}B, orange curve) is used as the optimal estimate of head tilt ($\tilde{H}_S$), given by the following:

@@

in which $\sigma_{HS}$ denotes the noise in the sensory signal, known to the observer, and $w_{HS}$ and $w_{HP}$ represent the relative weights of the sensory signal and the prior, respectively. Note that \eqnref{p1:eqn13} is equivalent to \eqnref{p1:eqn1} in \nameref{p1:sec:methods}. The variance of the posterior distribution in a single trial is given by the following:

@@

and is reflected by the width of the orange curve in \figref{p1:fig8}B. \figref{p1:fig8}D-F illustrates performance in multiple trials, in which the posterior distributions vary due to sensory noise ($\sigma_{HS}$), whereas the prior remains fixed. The variance of each posterior distribution is fixed and is given by \eqnref{p1:eqn16}.

@@

That the variance of the peak locations across multiple trials, $\sigma^2(\tilde{H}_S)$, is smaller can be shown by applying the rules of noise propagation to \eqnref{p1:eqn13}.

@@

which is equivalent to \eqnref{p1:eqn10} in Materials and Methods. Corresponding panels G-I (\figref{p1:fig8}) illustrate the distribution of the sensory signals for a given tilt angle (green-shaded curve), the prior distribution (blue-shaded curve), and the optimal estimates (orange-shaded curve), respectively. \figref{p1:fig8}I illustrates that the distribution of the optimal estimates across many trials has a lower variance than the posterior distribution in each single trial (\figref{p1:fig8}B), which follows from the comparison of \eqnref{p1:eqn16,p1:eqn17}, respectively.

### 2014-09-10 src/paper1/p1_intro.tex
Updated references
New siang command. Created custom style file.

    \caption{Schematic representation of the sensory integration model. Sensory signals, denoted by a hat symbol (\textasciicircum), are assumed to be calibrated accurately, but contaminated by Gaussian noise. Optimal estimates are denoted by a tilde (\textasciitilde). Body sensors, neck sensors, and otoliths provide information about orientation of body in space ($B_S$), head on body ($H_B$), and head in space ($H_S$), respectively. Neck signal ($\hat{H}_B$) is used for a reference frame transformation of otolith information into a body-in-space signal ($\hat{H}_S - \hat{H}_B = \hat{B}_{SI}$), and for a transformation of body-tilt information into a head-in-space signal ($\hat{B}_S + \hat{H}_B = \hat{H}_{SI}$). For an optimal estimate of body-in-space orientation, $\tilde{B}_S$ (SBT task), the model combines the body-sensor signal ($\hat{B}_S$, red pathway) with a reference-frame-transformed otolith signal ($\hat{B}_{SI}$, green pathway). Relative contributions of the two pathways ($w_{BD}$ and $w_{BI}$) depend on their relative precision (Eq. \ref{p1eqn2}). The scheme shows a symmetrical arrangement with two priors, but there is ample reason to believe that their effects are not identical. The simplest explanation of current and previous SBT data (see \nameref{p1:sec:methods}, \nameref{p1:sec:sbt_computation}) indicates that the associated prior in this task is uniform, which implies that $w_{BP}$ can be ignored. In the SVV task, an optimal estimate of head-in-space ($\tilde{H}_S$) is obtained by integration of otolith information ($\hat{H}_S$, green pathway), reference-frame-transformed information from body sensors ($\hat{H}_{SI}$, red pathway), and a significant contribution from prior information ($H_{SP}$, blue pathway). Relative weights are denoted by $w_{HD}$, $w_{HI}$, and $w_{HP}$, respectively. Estimate of line-in-space orientation is obtained by combining $\tilde{H}_S$ and estimates of eye-in-head ($\tilde{E}_H$) and line-on-eye ($\tilde{L}_E$) orientation. Noise variance in body sensors ($\sigma^2_{BS}$), neck sensors ($\sigma^2_{HB}$), otoliths ($\sigma^2_{HS}$), and width of prior ($\sigma^2_{HSP}$) defines their relative weights (see \nameref{p1:sec:methods}). Otolith noise may depend on tilt angle (Eq. \ref{p1:eqn11}). Note that the process of sensory integration, denoted here by summation of weighted sensory signals, is equivalent to multiplication of the underlying probability distributions (Eqs. \ref{p1:eqn2} and \ref{p1:eqn6} and \nameref{p1:sec:appendix}).}
    \label{p1:fig1}
\end{figure}

>>>>

    \caption{Schematic representation of the sensory integration model. Sensory signals, denoted by a hat symbol (\textasciicircum), are assumed to be calibrated accurately, but contaminated by Gaussian noise. Optimal estimates are denoted by a tilde (\textasciitilde). Body sensors, neck sensors, and otoliths provide information about orientation of body in space ($B_S$), head on body ($H_B$), and head in space ($H_S$), respectively. Neck signal ($\hat{H}_B$) is used for a reference frame transformation of otolith information into a body-in-space signal ($\hat{H}_S - \hat{H}_B = \hat{B}_{SI}$), and for a transformation of body-tilt information into a head-in-space signal ($\hat{B}_S + \hat{H}_B = \hat{H}_{SI}$). For an optimal estimate of body-in-space orientation, $\tilde{B}_S$ (SBT task), the model combines the body-sensor signal ($\hat{B}_S$, red pathway) with a reference-frame-transformed otolith signal ($\hat{B}_{SI}$, green pathway). Relative contributions of the two pathways ($w_{BD}$ and $w_{BI}$) depend on their relative precision (\eqnref{p1:eqn2}). The scheme shows a symmetrical arrangement with two priors, but there is ample reason to believe that their effects are not identical. The simplest explanation of current and previous SBT data (see \nameref{p1:sec:methods}, \nameref{p1:sec:sbt_computation}) indicates that the associated prior in this task is uniform, which implies that $w_{BP}$ can be ignored. In the SVV task, an optimal estimate of head-in-space ($\tilde{H}_S$) is obtained by integration of otolith information ($\hat{H}_S$, green pathway), reference-frame-transformed information from body sensors ($\hat{H}_{SI}$, red pathway), and a significant contribution from prior information ($H_{SP}$, blue pathway). Relative weights are denoted by $w_{HD}$, $w_{HI}$, and $w_{HP}$, respectively. Estimate of line-in-space orientation is obtained by combining $\tilde{H}_S$ and estimates of eye-in-head ($\tilde{E}_H$) and line-on-eye ($\tilde{L}_E$) orientation. Noise variance in body sensors ($\sigma^2_{BS}$), neck sensors ($\sigma^2_{HB}$), otoliths ($\sigma^2_{HS}$), and width of prior ($\sigma^2_{HSP}$) defines their relative weights (see \nameref{p1:sec:methods}). Otolith noise may depend on tilt angle (\eqnref{p1:eqn11}). Note that the process of sensory integration, denoted here by summation of weighted sensory signals, is equivalent to multiplication of the underlying probability distributions (\eqnref[ and ]{p1:eqn2,p1:eqn6} and \nameref{p1:sec:appendix}).}
    \label{p1:fig1}
\end{figure}
